Towards a Reinforcement Learning approach for Aggregate Computing

### 2021-11-28 paper-2021-woa-ac-ml.tex
minors: remove a comma in section III
grammars minors

\usepackage{tcolorbox}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}
\usepackage{xcolor}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{minted}
\usepackage{subcaption}
\usepackage[hidelinks, draft]{hyperref}
\usepackage[inline]{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{systeme}

@@

Initially, researchers had tried to achieve these properties in computing systems by miming natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to pursue the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours by a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different stuffs not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt w.r.t. the node movements rapidity.
%
Ideally, we would like to continue building our application using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
 to developers of defining the right collective specifications.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

\subsection{Computational Model}
%
For an aggreate program to yield collective adaptive behaviour, 
 continuous computation and communication is needed.
%
An individual atomic step 
 involving context acquisition,
 computation of the program,
 and neighbourhood-based communication 
 is called a \round{}.
%
In detail, a \round{} run at some device consists of:
\begin{enumerate}
  \item \emph{context acquisition}: the device collects information from the sensors, 
  the messages from neighbours
  (including the device itself---to model state);
  \item \emph{program evaluation}: the aggregate program is evaluated against the acquired context,
  yielding an \export{}, namely the message to be sent to neighbours;
  \item \emph{export sharing}: the \export{} is sent to the neighbours;
  \item \emph{actuations}: the \export{} also includes information that can be used to drive actuators in the local node.
\end{enumerate}

Rounds execution is completely asynchronous, so there is no global clock or barrier to coordinate the aggregate. 
%
 This makes it possible to scale easily with the number of nodes. 
%
The combination of the logic of the aggregate program
 and 
 such a collective, proactive, and periodical execution
 can promote the pursuing of collective adaptive behaviour.

@@

As a first step towards a complete integration between AC and RL, we haved decided to set up an Independent RL learning technique, 
 by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.

>>>>

\usepackage{tcolorbox}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}
\usepackage{xcolor}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{minted}
\usepackage{subcaption}
\usepackage[draft]{hyperref}
\usepackage[inline]{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{systeme}

@@

Initially, researchers had tried to achieve these properties in computing systems by miming natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to pursue the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different stuff not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt w.r.t. the node movements rapidity.
%
Ideally, we would like to continue building our application using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
 to developers of defining the right collective specifications.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

\subsection{Computational Model}
%
For an aggregate program to yield collective adaptive behaviour, 
 continuous computation and communication are needed.
%
An individual atomic step 
 involving context acquisition,
 computation of the program,
 and neighbourhood-based communication 
 is called a \round{}.
%
In detail, a \round{} run at some device consists of:
\begin{enumerate}
  \item \emph{context acquisition}: the device collects information from the sensors, 
  the messages from neighbours
  (including the device itself---to model state);
  \item \emph{program evaluation}: the aggregate program is evaluated against the acquired context,
  yielding an \export{}, namely the message to be sent to neighbours;
  \item \emph{export sharing}: the \export{} is sent to the neighbours;
  \item \emph{actuations}: the \export{} also includes information that can be used to drive actuators in the local node.
\end{enumerate}

Rounds execution is completely asynchronous, so there is no global clock or barrier to coordinate the aggregate. 
%
 This makes it possible to scale easily with the number of nodes. 
%
The combination of the logic of the aggregate program
 and 
 such a collective, proactive, and periodical execution
 can promote the pursuit of collective adaptive behaviour.

@@

As a first step towards a complete integration between AC and RL, we have decided to set up an Independent RL learning technique, 
 by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.
